Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-10-14 FreeLogic/2018-JAR/article.tex
added

\begin{defn}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subset D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  felshood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{defn}

\begin{defn}[Evaluation Function for $\FFOL$]
  The \emph{value} $\| s \|^{M,g}$ of a term or formula $s\in\FFOL$
  in a model $M=\langle D, \ex, I, \star\rangle$ under assignment $g$ 
  defined in the following way:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}

@@

We assume familiarity with $\lambda$-conversion (e.g. \emph{$\alpha$-renaming} and
\emph{$\beta\eta$-reduction}) and \emph{$\beta\eta$-normal forms}~\cite{barendregt13book}.

@@

\begin{defn}[Henkin model $\map{M}$ for $\FFOL$ model $M$]\label{def:hm}
Given a $\FFOL$ model $M = \langle D, \ex, I, \star\rangle$. The Henkin
model $\map{M} = \hol{\langle \{D_\alpha\}_{\alpha \in {T}},I \rangle}$ for
$M$ is defined as follows: 
\begin{itemize}
\item $\hol{D_i} = D$
\item $\hol{D_o} = \{\True,\False\}$
\item $\hol{D_{\alpha\ar\beta}}$ are
chosen as (not necessarily full) collections of functions from
$\hol{D_\alpha}$ to $\hol{D_\beta}$. Remember, however, that the
choice of $\hol{D_{\alpha\ar\beta}}$ must always ensure that 
the evaluation function 
 $\hol{\|\cdot\|^{\map{M},\map{g}}}$ below remains total, i.e. that
 all terms denote. In particular, it is required that $\hol{D_{i\ar
     o}}$ contains the element $\hol{I \ex_{i\ar
     o}}$ as characterized below.
\item The interpretation function $\hol{I}$ of $\map{M}$
  is chosen as follows:
\begin{itemize}
\item  $\hol{I \star_i} = \star \in \hol{D_i}$
\item  For all
    $\hol{d} \in \hol{D_i}$ we have: $(\hol{I \ex_{i\ar o}})(\hol{d}) =
    \hol{T}$  if and only if $\hol{d} \in E$. Note that this implies $(\hol{I \ex_{i\ar o}})(\hol{\star_i}) = \hol{F}$.
\item \sloppy For all $\hol{f} = \map{f} \in \map{F}$ we
       have: $(\hol{I} \hol{f})(\hol{d^1},\ldots,\hol{d^n}) = (I f)(\hol{d^1},\ldots,\hol{d^n})$ for
       all $\hol{d^i}\in \hol{D_i}$ ($i=1,\ldots,n$ and
       $n\geq 0$).
\item \sloppy For all $\hol{p} = \map{p} \in \map{P}$ we
       have: $(\hol{I} \hol{p})(\hol{d^1},\ldots,\hol{d^n}) = \hol{T}$
       if and only if $(\hol{d^1},\ldots,\hol{d^n}) \in (I p)$ for
       all $\hol{d^i}\in \hol{D_i}$ ($i=1,\ldots,n$ and
       $n\geq 0$).
\item For all other constants $\hol{c_\alpha}$, choose $\hol{I
    c_\alpha} \in \hol{D_\alpha}$ arbitrary.\footnote{In fact, it may be safely assumed that there are
  no other constant symbols given in a HOL signature, except for the symbols in
  $\map{F}$ and $\map{P}$, the symbols $\hol{\ex_{i\ar o}}$  and $\hol{\star_{i}}$ and the logical connectives.}
\end{itemize}
\end{itemize}

@@

\begin{table}[htp] \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set I} \\
%\hline
\\
 $S_{i}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y)$ \\
 $E_{i}$ & $\ex(x \comp y) \limp (\ex x \wedge \ex y \wedge (\exists z\lambdot z \comp
         z \cong z \wedge x \comp z \cong x \wedge z \comp y \cong y))$ \\
 $A_{i}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{i}$ & $\forall y\lambdot \exists i\lambdot  \id i \wedge i \comp y \cong y$ \\
 $D_{i}$ & $\forall x\lambdot \exists j\lambdot  \id j \wedge x \comp j \cong x$  \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set II} \\
%\hline
\\
$S_{ii}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{ii}$ & $\ex(x \comp y) \limp (\ex x \wedge \ex y \wedge (\exists z\lambdot z \comp
         z \cong z \wedge x \comp z \cong x \wedge z \comp y \cong y))$ \\
 $A_{ii}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{ii}$ & $\ex y \imp (\id(\cod y) \wedge (\cod y) \comp y \cong y)$ \\
 $D_{ii}$ & $\ex x \imp (\id(\dom x) \wedge x \comp (\dom x) \cong x)$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set III} \\
%\hline
\\
$S_{iii}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{iii}$ & $\ex(x \comp y) \limp (\dom x \cong \cod y \wedge \ex (\cod y)))$ \\
 $A_{iii}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{iii}$ & $\ex y \imp (\id(\cod y) \wedge (\cod y) \comp y \cong y)$ \\
 $D_{iii}$ & $\ex x \imp (\id(\dom x) \wedge x \comp (\dom x) \cong x)$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set IV} \\
%\hline
\\
$S_{iv}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{iv}$ & $\ex(x \comp y) \biimp (\dom x \cong \cod y \wedge \ex (\cod y)))$ \\
 $A_{iv}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{iv}$ & $(\cod y) \comp y \cong y$ \\
 $D_{iv}$ & $x \comp (\dom x) \cong x$  \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set V (Scott 79, \cite{Scott79})} \\
%\hline
\\
$S1$ & $\ex (\dom x) \imp \ex x$ \\
$S2$ & $\ex (\cod y) \imp \ex y$ \\
$S3$ & $\ex(x \comp y) \biimp \dom x \simeq \cod y$ \\
$S4$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
$S5$ & $(\cod y) \comp y \cong y$ \\
$S6$ & $x \comp (\dom x) \cong x$  \\
\\
\hline
\end{tabular}
\caption{Stepwise evolution of Scott's \cite{Scott79} axiom
  system for category theory from partial monoids. The axiom names are
  motivated as follows: 
  $S$ stands for strictness, $E$ for existence, $A$ for associativity, $C$ for
  codomain, $D$ for Domain. The free variables $x$, $y$, $z$ range over
  the raw domain $D$. The quantifiers in Axioms Sets I and II are
  free logic quantifiers, that is, they range over the domain $E$ of
  existing objects. \label{axioms-sets-1}}
\end{table}

>>>>

\begin{defn}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subset D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  falsehood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{defn}

\begin{defn}[Evaluation Function for $\FFOL$] \sloppy
  The \emph{value} $\| s \|^{M,g}$ of a term or formula $s\in\FFOL$
  in a model $M=\langle D, \ex, I, \star\rangle$ under assignment $g$ 
  defined in the following way:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}

@@

\sloppy We assume familiarity with $\lambda$-conversion (e.g. \emph{$\alpha$-renaming} and
\emph{$\beta\eta$-reduction}) and \emph{$\beta\eta$-normal forms}~\cite{barendregt13book}.

@@

\begin{defn}[Henkin model $\map{M}$ for $\FFOL$ model
  $M$]\label{def:hm} \sloppy
Given a $\FFOL$ model $M = \langle D, \ex, I, \star\rangle$. The Henkin
model $\map{M} = \hol{\langle \{D_\alpha\}_{\alpha \in {T}},I \rangle}$ for
$M$ is defined as follows: 
\begin{itemize}
\item $\hol{D_i} = D$
\item $\hol{D_o} = \{\True,\False\}$
\item $\hol{D_{\alpha\ar\beta}}$ are
chosen as (not necessarily full) collections of functions from
$\hol{D_\alpha}$ to $\hol{D_\beta}$. Remember, however, that the
choice of $\hol{D_{\alpha\ar\beta}}$ must always ensure that 
the evaluation function 
 $\hol{\|\cdot\|^{\map{M},\map{g}}}$ below remains total, i.e. that
 all terms denote. In particular, it is required that $\hol{D_{i\ar
     o}}$ contains the element $\hol{I \ex_{i\ar
     o}}$ as characterized below.
\item The interpretation function $\hol{I}$ of $\map{M}$
  is chosen as follows:
\begin{itemize}
\item  $\hol{I \star_i} = \star \in \hol{D_i}$
\item  For all
    $\hol{d} \in \hol{D_i}$ we have: $(\hol{I \ex_{i\ar o}})(\hol{d}) =
    \hol{T}$  if and only if $\hol{d} \in E$. Note that this implies $(\hol{I \ex_{i\ar o}})(\hol{\star_i}) = \hol{F}$.
\item \sloppy For all $\hol{f} = \map{f} \in \map{F}$ we
       have: $(\hol{I} \hol{f})(\hol{d^1},\ldots,\hol{d^n}) = (I f)(\hol{d^1},\ldots,\hol{d^n})$ for
       all $\hol{d^i}\in \hol{D_i}$ ($i=1,\ldots,n$ and
       $n\geq 0$).
\item \sloppy For all $\hol{p} = \map{p} \in \map{P}$ we
       have: $(\hol{I} \hol{p})(\hol{d^1},\ldots,\hol{d^n}) = \hol{T}$
       if and only if $(\hol{d^1},\ldots,\hol{d^n}) \in (I p)$ for
       all $\hol{d^i}\in \hol{D_i}$ ($i=1,\ldots,n$ and
       $n\geq 0$).
\item For all other constants $\hol{c_\alpha}$, choose $\hol{I
    c_\alpha} \in \hol{D_\alpha}$ arbitrary.\footnote{In fact, it may be safely assumed that there are
  no other constant symbols given in a HOL signature, except for the symbols in
  $\map{F}$ and $\map{P}$, the symbols $\hol{\ex_{i\ar o}}$  and $\hol{\star_{i}}$ and the logical connectives.}
\end{itemize}
\end{itemize}

@@

\begin{table}[htp] \centering \normalsize
\begin{tabular}{ll}
\hline
\\[-.8em]
\multicolumn{2}{c}{Axioms Set I} \\
%\hline
\\
 $S_{i}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y)$ \\
 $E_{i}$ & $\ex(x \comp y) \limp (\ex x \wedge \ex y \wedge (\exists z\lambdot z \comp
         z \cong z \wedge x \comp z \cong x \wedge z \comp y \cong y))$ \\
 $A_{i}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{i}$ & $\forall y\lambdot \exists i\lambdot  \id i \wedge i \comp y \cong y$ \\
 $D_{i}$ & $\forall x\lambdot \exists j\lambdot  \id j \wedge x \comp j \cong x$  \\
\\
\hline
\\[-.8em]
\multicolumn{2}{c}{Axioms Set II} \\
%\hline
\\
$S_{ii}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{ii}$ & $\ex(x \comp y) \limp (\ex x \wedge \ex y \wedge (\exists z\lambdot z \comp
         z \cong z \wedge x \comp z \cong x \wedge z \comp y \cong y))$ \\
 $A_{ii}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{ii}$ & $\ex y \imp (\id(\cod y) \wedge (\cod y) \comp y \cong y)$ \\
 $D_{ii}$ & $\ex x \imp (\id(\dom x) \wedge x \comp (\dom x) \cong x)$ \\
\\
\hline
\\[-.8em]
\multicolumn{2}{c}{Axioms Set III} \\
%\hline
\\
$S_{iii}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{iii}$ & $\ex(x \comp y) \limp (\dom x \cong \cod y \wedge \ex (\cod y)))$ \\
 $A_{iii}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{iii}$ & $\ex y \imp (\id(\cod y) \wedge (\cod y) \comp y \cong y)$ \\
 $D_{iii}$ & $\ex x \imp (\id(\dom x) \wedge x \comp (\dom x) \cong x)$ \\
\\
\hline
\\[-.8em]
\multicolumn{2}{c}{Axioms Set IV} \\
%\hline
\\
$S_{iv}$ & $\ex(x \comp y) \imp (\ex x \wedge \ex y) \wedge (\ex (\dom x) \imp \ex
        x) \wedge (\ex (\cod y) \imp \ex
        y)$ \\
 $E_{iv}$ & $\ex(x \comp y) \biimp (\dom x \cong \cod y \wedge \ex (\cod y)))$ \\
 $A_{iv}$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
 $C_{iv}$ & $(\cod y) \comp y \cong y$ \\
 $D_{iv}$ & $x \comp (\dom x) \cong x$  \\
\\
\hline
\\[-.8em]
\multicolumn{2}{c}{Axioms Set V (Scott 79, \cite{Scott79})} \\
%\hline
\\
$S1$ & $\ex (\dom x) \imp \ex x$ \\
$S2$ & $\ex (\cod y) \imp \ex y$ \\
$S3$ & $\ex(x \comp y) \biimp \dom x \simeq \cod y$ \\
$S4$ & $x\comp (y \comp z) \cong (x \comp y) \comp z$ \\
$S5$ & $(\cod y) \comp y \cong y$ \\
$S6$ & $x \comp (\dom x) \cong x$  \\
\\
\hline
\end{tabular}
\caption{Stepwise evolution of Scott's \cite{Scott79} axiom
  system for category theory from partial monoids. The axiom names are
  motivated as follows: 
  $S$ stands for strictness, $E$ for existence, $A$ for associativity, $C$ for
  codomain, $D$ for Domain. The free variables $x$, $y$, $z$ range over
  the raw domain $D$. The quantifiers in Axioms Sets I and II are
  free logic quantifiers, that is, they range over the domain $E$ of
  existing objects. \label{axioms-sets-1}}
\end{table}
